---
title: "03-lab3.qmd"
format: html
editor: visual
---

Packages required during R Session:

```{r}
#| echo: false
library("IRanges")
library("GenomicRanges")
library("Rsamtools")
library("GenomicAlignments")
library("Gviz")
```

## Question 1:

line 1: Load string package to manipulate string characters

line 2: Read in log file into vector that can be parsed.

line 3: Create a tibble to tabulate the statistics found at the end of the document.

```{r}
# load library to use stringr
library(stringr)

# read in file 
summary_stats <- readLines("outputs/flye_output/flye.log")

assembly_stats <- data.frame(
  "Metric" = c("Total assembly lenght",
               "N50", 
               "Num of contigs",
               "Largest contig",
               "Coverage"),
  "Values" = as.numeric(c(str_extract(summary_stats[13362], "10207250")),
                        str_extract(summary_stats[13364], "128892"),
                        str_extract(summary_stats[13363], "430"),
                        str_extract(summary_stats[13365], "85335"),
                        str_extract(summary_stats[13367], "6")
                        )
)
```

- Q: Based on that, would you expect to have a complete genome? Would you say that the coverage is adequate?

  - No, as the mean coverage is low at 6x indicating each assembled base is covered by 6 read bases on average.

## Question 2:

The `BUSCO` reports should be found under `outputs/busco_output`. Open the text file with lineage specific results, and answer the following:

- What were the `BUSCO` scores? Given the coverage do you think they are good (2 marks)?

  - The busco score is **C**omplete:42.8% \[**S**ingle-copy:42.7%, **D**uplicated:0.1%\], **F**ragmented:0.6%, **M**issing:56.6.

  - Based on the low mean coverage of 6x the low BUSCO completeness of 42.8% is not good as almost 56.6% of the genome is missing.

- What was the N50 compared to the `flye` calculated one (1 mark)?

  - BUSCO calculated N50 is relatively close to the flye calculated one. BUSCO seems rounded with value as 28 kbp, while flye is the whole number 28892 bps.

- What was the total assembly length (1 mark)?

  - total assembly length is 10038806 bps.

## Question 3:

Open the `quast` report:

- What was the genome fraction how does it compare to `BUSCO` (2 marks)?

  - Genome fraction is 4.924% indicates the percentage of the reference genome that is covered by assembly, while Busco gives insight into the expected amount of conserved genes found in the assembly. They provide different context into the assembly.

- What was the duplication ratio (1 mark)?

  - 1.017

- What was the level of misassemblies and what is the length of the misassembled contigs? What does it tell you about how well the reference genome represents the genomic diversity of the species *S. cerevisiae* (3 marks)?

  - The level of misassemblies is 2, and the length of the misassembled contigs is 102 152 bp.
  - A misassembly indicates a structural inconsistancy between the assembly and the reference alignment, but it does not necessarily mean the assembly is correct.
  - These values alone do not provide sufficient information to interpret genomic diversity S*. cerevisiae* as the reason why there is a misassembly needs to be further investigated between technical error or true biological insight.

## Question 4:

```{r}
library(Biostrings)

# Create path to file 
chr_path <- "data/S288C_UCSC.fna"

# Read in .fna file 
chr <- readLines(chr_path)

# extract chromosome 1 from file 
header_lines <- grep(">", chr)

# Safety Check: Yeast has 16 nuclear chromosomes + 1 mitochondrial
if(length(header_lines) == 17) {
    if (chr[header_lines[1]] == ">chrI") {
 
     # chr1 contains only chromsome 1 sequence 
     chrI <- chr[header_lines[1]:(header_lines[2] - 1)]
     
     writeLines(chrI, "data/R64_chr1.fna")
     
     message("SUCCESS: chrI has been found and extracted.")
   }
 } else {
   stop("Error: Chromosome count mismatch! Check your input file.")
 }
```

## Question 5:

The `echo: false` option disables the printing of code (only output is displayed).

```{r}
#| echo: false
# BAM file path
bam_file <- "outputs/alignment.sorted.bam"

# Open connection to BAM file
bam <- BamFile(bam_file)

# Read all alignments
alns <- readGAlignments(bam)

# Convert to GRanges object
gr_alns <- granges(alns)

aln_track <- AnnotationTrack(gr_alns, name = "Contigs", genome = "sacCer3", chromosome = "chrI")

# Add genome axis for scale
axis_track <- GenomeAxisTrack()

# Plot to visualize the alignment
plotTracks(list(axis_track, aln_track),
           from = min(start(gr_alns)),
           to = max(end(gr_alns)),
           main = "Contig Alignments to Chromosome 1")
```

- How many contigs did you have aligning to chromosome 1 (1 mark)?

  *(Following bash code from chatgpt)*

  ```{bash}
  samtools view -F 4 outputs/alignment.sorted.ba | cut -f1 | sort -u
  ```

  - Returned 7 unique contigs aligned to Chromosome I: \
    contig_100 contig_20 contig_309 contig_321 contig_333 contig_380 contig_68

- Using the functions `range()` and `ranges()` try and estimate how much of chromosome 1 does your contigs cover, is it in line with the alignment results reported by `QUAST` (2 marks)?

  ```{r}
  # Find the start and end boundaries of all mapped contigs combined
  range(gr_alns)
  # Overall on chromosome 1, the contigs spans approximately 1 - 222787 bp. 

  # Peek at the individual contig boundaries
  ranges(gr_alns)
  # Individually on chromosome 1, the contigs cover roughly 17016 bp. 
  ```

  - Compared to QUAST, the BAM alignment results are not in line with report.

## Question 6:

```{r}
# Path to file consensus.fasta
consensus_path <- "outputs/medaka_round2/consensus.fasta"

# read in contigs
contigsChrI <- readDNAStringSet(consensus_path)

# filter chromsome 1 file for longest contig
longest_contig <- contigsChrI[which.max(width(contigsChrI))]

# write and output longest contig 
writeXStringSet(
  longest_contig,
  "outputs/longest_contig.fasta"
)

```

## Question 7:

Convert SAM file into indexed BAM file for visualization

```{bash}
samtools view -bS outputs/cds_aln.sam > outputs/cds_aln.bam
samtools sort outputs/cds_aln.bam -o outputs/cds_aln.sorted.bam
samtools index outputs/cds_aln.sorted.bam
```

## Question 8:

```{r}
#| echo: false
# BAM file path
bam_file8 <- "outputs/cds_aln.sorted.bam"

# Open connection to BAM file
bam8 <- BamFile(bam_file8)

# Read all alignments
alns8 <- readGAlignments(bam8)

# Convert to GRanges object
gr_alns8 <- granges(alns8)

# Used as alignment is not named according to UCSC browser convention
options(ucscChromosomeNames = FALSE)

aln_track8 <- AnnotationTrack(gr_alns8, name = "CDS Alignment", genome = "chrI", chromosome = "contig_23")

# Add genome axis for scale
axis_track8 <- GenomeAxisTrack()

# Plot to visualize the alignment
plotTracks(list(axis_track8, aln_track8),
           from = min(start(gr_alns8)),
           to = max(end(gr_alns8)),
           main = "Contig Alignments to Longest Contig")

# Set back to TRUE, to ensre environment code is not impacted. 
options(ucscChromosomeNames = TRUE)
```
